---
title: "OPTT Analysis"
author: Oliver Reece
format: pdf
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
#install.packages("tidyquant") #for ticker price data
library("tidyquant") 
#install.packages("lubridate") #for date manipulation
library("lubridate")
#install.packages("tidyverse")
library("tidyverse")
#install.packages("dplyr")
library("dplyr")
#install.packages("zoo") #for time-series analysis
library("zoo")
#install.packages("httr") #for sending API requests
library("httr")
#install.packages("jsonlite") #convert json to R text (for parsing API return data)
library("jsonlite")
#install.packages("newsanchor") #easiest way to query NEWSAPI
library("newsanchor")
```

```{r}
#api keys
finnhub_key <- Sys.getenv("FINNHUB_API_KEY")
news_api_key <- Sys.getenv("NEWS_API_KEY")
```

```{r}
#pull OPTT ticker data for all time since going public.
#date format: YYYY-MM-DD

optt <- tq_get("OPTT",
               get = "stock.prices",
               from = "2007-04-25", #went public on NASDAQ April 25th 2007
               to = "2026-06-21") #at time of download, update as needed
```

```{r}
#preview data
head(optt)
```

```{r}
#summary table
summary(optt)
```

```{r}
#tidy datatable

#rename variables to be more 
optt_tidy <- optt |> 
  rename(open_price = open,
         high_price = high,
         low_price = low,
         close_price = close,
         Volume = volume,
         Adjusted = adjusted) |> 

#break date into day, month, year columns
  separate(date, into = c("Year", "Month", "Day"), sep = "-", convert = TRUE) |> 
  
#add month name column
  mutate(month_name = month.name[Month]) |> 
  
#place month name column after month number
  relocate(month_name, .after = Month) |> 

#create new variables
  mutate(
    #intra-day volatility
    idv = (high_price-low_price),
    #previous close price
    prev_close = lag(close_price),
    #absolute maximum upward movement
    max_up = (abs(high_price - prev_close)),
    #absolute maximum downward movement
    max_down = (abs(low_price - prev_close)),
    #true range:
    #use pmax (parallel max) to take max for each day --> eliminates need for if/else + max()
    true_range = pmax(idv, max_up, max_down, na.rm = TRUE),
    #average true range:
    #use rollmean (zoo package) to look at a sliding window
    ATR = rollmean(true_range, k = 14, fill = NA, align = "right")
  )
```

```{r}
#preview tidy data
head(optt_tidy)
```

```{r}
#view summary of tidy data
summary(optt_tidy)
```

## Sentiment Analysis Scoring 

```{r}
#finnhub Sentiment News Scoring

#connect to finnhub
finnhub_url <- paste0(
  "https://finnhub.io/api/v1/company-news?symbol=OPTT&from=2026-05-21&to=2026-06-21&token=", finnhub_key)

#process data using httr and jsonlite
finnhub_raw_data <- GET(finnhub_url)
finnhub_parsed <- fromJSON(content(finnhub_raw_data, "text"))

#preview parsed data
head(finnhub_parsed)
str(finnhub_parsed)

#clean parsed data into a usable format
finnhub_tidy <- finnhub_parsed |> 
  mutate(
    Date = as_date(as_datetime(datetime)),
    Text = headline
  ) |> 
  rename(Source = source) |> 
  select(Date, Text, Source, url)

#only 1 article
```

```{r}
# News API Sentiment Scoring 

#connect to NEWSAPI using newsanchor
news_api_result <- get_everything(
  query = '"Ocean Power Technologies" OR "OPTT stock" OR "NASDAQ:OPTT"',
  from = "2026-05-21",
  to = "2026-06-21",
  api_key = news_api_key
)

#check to see if we got data
news_api_result$metadata$total_results   
news_api_result$metadata$status_code     
news_api_result$results_df  

#3 articles
```

After querying both NewsAPI and Finnhub, we returned a total of 4 relevant articles. 
Due to concerns of limited data and subsequent biases, I will not be using sentiment analysis 
scores within my models. 

Instead, the analysis will focus on price-derived technical indicators alongside trading volume. 
These variables are derived directly from empirical trading data, avoiding the sample-size and 
bias concerns associated with the limited sentiment data.














