---
title: "OPTT Analysis"
author: Oliver Reece
format: pdf
execute:
  echo: false
  warning: false
  message: false
  output: false
---

```{r}
#install.packages("tidyquant") #for ticker price data
library("tidyquant") 
#install.packages("lubridate") #for date manipulation
library("lubridate")
#install.packages("tidyverse")
library("tidyverse")
#install.packages("dplyr")
library("dplyr")
#install.packages("zoo") #for time-series analysis
library("zoo")
#install.packages("httr") #for sending API requests
library("httr")
#install.packages("jsonlite") #convert json to R text (for parsing API return data)
library("jsonlite")
#install.packages("newsanchor") #easiest way to query NEWSAPI
library("newsanchor")
#install.packages("knitr")
library("knitr")
#install.packages("TTR") #pre-built financial analysis indicators
library("TTR")
#install.packages("kableExtra") #used for editing kable tables
library(kableExtra)

```

```{r}
#api keys
finnhub_key <- Sys.getenv("FINNHUB_API_KEY")
news_api_key <- Sys.getenv("NEWS_API_KEY")
```

# Abstract

Founded in 1894, Ocean Power Technologies (OPTT) is a martime-focused company with an emphasis on
finding solutions to the need for more efficient/productive infrastructure. Their
services span defense and security, oil and gas procurement, and off-shore wind markets.

OPTT went public on the NASDAQ on the 25th of April, 2007. Initially, the stock 
exploded, with a closing price high of $3,652.00 on November 12th, 2007. Ever since, 
the stock has been on the decline, falling under $5 per share in early 2019. Crossing
that $5 threshold means the stock is now classified as a penny stock. Ocean Power
Technologies hosts a market cap of $74.6 Million (at the time of writing), thus 
designating the company as a micro-cap company. 

Since approximately 2012, OPTT has suffered in many ways that have led to investor
dis-trust. The three main areas of concern are the high risk designation due to 
high volatility, share dilution, and multi-million dollar operating losses. For
example, between 2012 and 2026, the number of shares (approximately) increased from
5 million to 180 million while total operating losses during that time exceeded
$300 Million.

That being said, OPTT is not entirely out of the market. On January 6th, 2026 
OPTT announced they had been awarded a contract by the U.S. Department of Homeland 
Security for the purpose of supplying the U.S. Coast Guard with their new 
PowerBuoy systems. The contract was valued at $6.5 Million. This contract was not
only a lifeline for OPTT but marked the transition from a sustainability focused 
private company into what is now essentially a revenue focused defense contractor. 

In this project, I use statistical modelling and financial metrics to 
analyse OPTT's performance and risk overtime. From there I aim to make inferences 
about whether thus far the contract has begun to aid OPTT recover. I then test 
my inferences using machine learning algorithms to predict future stock prices. 

I will be using OLS regression models for performance analysis, using the log of 
Volume as the dependent variable. I am using volume because in contrast to the
other metrics I can choose from, Volume is able to differentiate between direction
and attention in the market. Volume focuses on how many shares are bought or sold, 
meaning it is a good measure of general activity surrounding the stock itself as
opposed to what the public think of the stock. I take the log of Volume due to
it being right-skewed with a median of 7,710 vs a mean of 1,974,392.

I will use a logistic regression model for risk analysis, using a "risk event" binary 
as the dependent variable. The risk event will be triggered when a significant 
price drop or sell-side volume spike occurs, or both. The models will be ordinal in nature, 
each with 3 "severity" tiers, defined as:

- Level 1 -> No Risk Event: Price Drop < 3% and Volume Multiple < 1.25
- Level 2 -> Moderate Risk Event: Price Drop between 3%-7% (inclusive) or Volume Multiple between 1.25-2 (inclusive) AND neither metric surpasses Level 3 thresholds.
- Level 3 -> High Risk Event: Price Drop > 7% or Volume Multiple > 2

I will then move into predictive modelling, using Random Forest, XGBoost, and LSTM Networks to 
predict future performance. 

For the LSTM Network, the adjusted closing price will be the 
outcome variable. I am using adjusted closing over opening or closing price due
to the notion that an ML algorithm will look at fluctuations between day as catastrophic
economic events if the drop is significant enough from one day to another. It is 
also of note that if that drop is caused by a volatility spike, the adjusted closing
price will catch it. Moreover, the adjusted closing price would back-trace historical 
data which would prevent the model from falling for false inter-day price gaps.

For the Random Forest and XGBoost algorithms, it does not make sense to predict 
price as a raw dollar metric due to the lack of trend-awareness that these
algorithms possess. Both algorithms function by breaking the data down into rectangular 
cross sections based on real historic data. Thus, these models will never predict
a future dollar price that exceeds the upper and lower bounds of the data - a 
reality that does not exist in the stock market. As a solution, I will predict 
returns using a calculated percent change metric as the outcome variable. This
percent change will be calculated using adjusted closing price

To conclude, I will compare the effectiveness of the 3 ML models using error scores
to determine which stand-alone algorithm performed best.  

Why does any of this matter? Firstly, I enjoy machine learning and investing, 
so this was a natural project for me to take on. Secondly, it is an interesting
question to ask - can traditional ML handle the noise of penny stocks? Currently, 
analysts rely on hybrid algorithms that incorporate classification models, and 
sentiment analysis. In most cases, LSTM's tend to confuse the noise of unexpected 
price spikes that are overly common in penny stocks for true trends, and I want to
see the comparison myself. 

# The Data

In order to take on this project, I utilise a variety of methods to procure data.
The main pathway is the `tidyquant` package from which I am able to download 
historic data about the desired ticker. `tidyquant` primarily leans on Yahoo
Finance for key metrics, but also pulls from several other databases such as 
the Federal Reserve Economic Data (FRED).

I also make queries to NewsAPI and Finnhub, two APIs that allow me to scrape for
media related to my chosen ticker in order to perform sentiment analysis in 
conjunction with the machine learning models. Unfortunately, there is limited 
results for Ocean Power Technologies and as such I do not continue down this path.

Using `tidyquant`, I retrieve data from April 25th, 2007 (IPO) until June 21st, 
2026 which is the date of writing, although the data becomes usable after row 34 
due to the staggered system in which some calculated metrics such as MACD work. 
That being said, the complete dataset boasts 4,819 observations across 17 
variables (22 when including ticker name, year, month, month name, and day).

Of these 17 variables, some notable measures include:

- **`open_price`**: Stock price at open on a given trading day
- **`close_price`**: Stock price at close on a given trading day
- **`Volume`**: Number of shares traded (bought or sold) during a given trading day
- **`true_range`**: Total price volatility by finding the greatest distance between current high, current low and previous closing price
- **`ATR`**: Moving average of true range over a 14-day period
- **`rsi`**: Relative Strength Index: Measures magnitude and speed of price fluctuations on a scale of 0-100 (momentum indicator)
- **`macd_hist`**: Measures if the stock's momentum is speeding up or slowing down
- **`obv`**: On-Balance Volume: Relationship metric between volume and price used to spot divergences

To illustrate the structured dataset, rows 34–39 are displayed below. Because 
several indicators rely on rolling or exponential moving-average calculations of 
varying lengths (e.g., a 14-day window for ATR and RSI, and a 26-day EMA for the 
MACD line, followed by a further 9-day EMA for the MACD signal line), the initial 
rows of the dataset default to NA values until each indicator has accumulated 
sufficient historical data to produce a valid output. The MACD signal line, in 
particular, does not return numeric values until row 34, as it requires the MACD 
line itself to first be calculated. Rows 34–39 are therefore displayed to ensure 
all indicators, including MACD and its derived histogram, are represented with 
complete, numeric, and non-NA data points.

```{r}
#pull OPTT ticker data for all time since going public.
#date format: YYYY-MM-DD

optt <- tq_get("OPTT",
               get = "stock.prices",
               from = "2007-04-25", #went public on NASDAQ April 25th 2007
               to = "2026-06-21") #at time of download, update as needed
```

```{r}
#preview data
head(optt)
```

```{r}
#summary table
summary(optt)
```

```{r}
#tidy datatable

#rename variables to be more 
optt_tidy <- optt |> 
  rename(open_price = open,
         high_price = high,
         low_price = low,
         close_price = close,
         Volume = volume,
         Adjusted = adjusted) |> 

#break date into day, month, year columns
  separate(date, into = c("Year", "Month", "Day"), sep = "-", convert = TRUE) |> 
  
#add month name column
  mutate(month_name = month.name[Month]) |> 
  
#place month name column after month number
  relocate(month_name, .after = Month) |> 

#create new variables
  mutate(
    #intra-day volatility
    idv = (high_price-low_price),
    #previous close price
    prev_close = lag(close_price),
    #absolute maximum upward movement
    max_up = (abs(high_price - prev_close)),
    #absolute maximum downward movement
    max_down = (abs(low_price - prev_close)),
    #true range:
    #use pmax (parallel max) to take max for each day --> eliminates need for if/else + max()
    true_range = pmax(idv, max_up, max_down, na.rm = TRUE),
    #average true range:
    #use rollmean (zoo package) to look at a sliding window
    ATR = rollmean(true_range, k = 14, fill = NA, align = "right"),
    #RSI - Relative Strength Index - measures momentum (magnitude and speed of price fluctuations, 0-100)
    rsi = RSI(close_price, n = 14),
    #MACD - Moving Average Convergence Divergence
    #helps identify whether stock has sustained movement or just noisy movement.
    macd_trendline = MACD(close_price, nFast = 12, nSlow = 26, nSig = 9)[, "macd"],
    macd_signal = MACD(close_price, nFast = 12, nSlow = 26, nSig = 9)[, "signal"],
    #the distance between macd trend and signal is macd histogram
    #acts as a measure of how strong the bullish or bearish momentum is 
    macd_hist = (macd_trendline - macd_signal),
    #On Balance Volume (OBV):
    #essentially functions as a validity check of the price trend. 
    obv = OBV(close_price, Volume)
  )
```

```{r}
#preview tidy data
head(optt_tidy)
```

```{r}
#view summary of tidy data
summary(optt_tidy)
```

```{r}
#| output: true

#professional table output using kable:
optt_tidy |> 
  select(open_price,close_price,Volume,true_range,ATR, rsi, macd_hist, obv) |> 
  slice(34:39) |> 
  kable() |> 
  kable_styling(font_size = 12)
```

# Sentiment Analysis

```{r}
#finnhub Sentiment News Scoring

#connect to finnhub
finnhub_url <- paste0(
  "https://finnhub.io/api/v1/company-news?symbol=OPTT&from=2026-05-21&to=2026-06-21&token=", finnhub_key)

#process data using httr and jsonlite
finnhub_raw_data <- GET(finnhub_url)
finnhub_parsed <- fromJSON(content(finnhub_raw_data, "text"))

#preview parsed data
head(finnhub_parsed)
str(finnhub_parsed)

#clean parsed data into a usable format
finnhub_tidy <- finnhub_parsed |> 
  mutate(
    Date = as_date(as_datetime(datetime)),
    Text = headline
  ) |> 
  rename(Source = source) |> 
  select(Date, Text, Source, url)

#only 1 article
```

```{r}
# News API Sentiment Scoring 

#connect to NEWSAPI using newsanchor
news_api_result <- get_everything(
  query = '"Ocean Power Technologies" OR "OPTT stock" OR "NASDAQ:OPTT"',
  from = "2026-05-21",
  to = "2026-06-21",
  api_key = news_api_key
)

#check to see if we got data
news_api_result$metadata$total_results   
news_api_result$metadata$status_code     
news_api_result$results_df  

#3 articles
```

-placeholder for data table with all 4 articles represented and brief discussion of them-

After querying both NewsAPI and Finnhub, we returned a total of 4 relevant articles. 
Due to concerns of limited data and subsequent biases, I will not be using sentiment analysis 
scores within my models. 

Instead, the analysis will focus on price-derived technical indicators alongside trading volume. 
These variables are derived directly from empirical trading data, avoiding the sample-size and 
bias concerns associated with the limited sentiment data.

In order to maintain structural robustness via alternative data streams, the analysis 
will pivot to integrate macroeconomic corporate health metrics, specifically metrics
such as outstanding share dilution and order backlog growth.

# Alternate Data Procurement

-placeholder-

tq_get --> financials --> filter for type == income, unnest quarterly. 

problem --> quarterly means 4 data points per year, not a lot of info (data-sparcity bias)

solution --> forward stepwise function to fill for weekly / monthly data (outstanding shares)

solution --> linear interpolation to fill for weekly / monthly data (order backlog growth)


# Linear Regression Model

-placeholder for written explanation of model-

big question: does volatility relate to volume, in general, across the whole dataset

```{}
#regression model + output
ols <- lm(log(Volume) ~ rsi + macd_hist + shares_outstanding + order_backlog_growth, data = optt_model)
summary(ols)
```

```{}
#check residuals
plot(ols)
```









