---
title: "OPTT Analysis"
author: Oliver Reece
format: pdf
execute:
  echo: false
  warning: false
  message: false
  output: false
---

```{r}
#install.packages("tidyquant") #for ticker price data
library("tidyquant") 
#install.packages("lubridate") #for date manipulation
library("lubridate")
#install.packages("tidyverse")
library("tidyverse")
#install.packages("dplyr")
library("dplyr")
#install.packages("zoo") #for time-series analysis
library("zoo")
#install.packages("httr") #for sending API requests
library("httr")
#install.packages("jsonlite") #convert json to R text (for parsing API return data)
library("jsonlite")
#install.packages("newsanchor") #easiest way to query NEWSAPI
library("newsanchor")
#install.packages("knitr")
library("knitr")
```

```{r}
#api keys
finnhub_key <- Sys.getenv("FINNHUB_API_KEY")
news_api_key <- Sys.getenv("NEWS_API_KEY")
```

## Abstract

-placeholder-

- What is OPTT? Founded? 
- Talk about high risk, share dilution, operating loss
- What is their current market valuation? (penny, market cap)
- Talk about deal with US Gov.
- What do I aim to do originally? What changed? Why?
- What models? 
- So what? Looking into whether traditional ML can handle the noise of penny stocks

## The Data

-placeholder-

- Where does data come from?
- Time frame? 
- Num_observations?
- Key variables?
- Omitted Alt Data

To illustrate the structured dataset, rows 14–19 are displayed below. Because the 
ATR calculation utilises a 14-day sliding window, the initial 13 observations by
default result in NA values. Examining rows 14–19 ensures the inclusion of complete data points.

```{r}
#pull OPTT ticker data for all time since going public.
#date format: YYYY-MM-DD

optt <- tq_get("OPTT",
               get = "stock.prices",
               from = "2007-04-25", #went public on NASDAQ April 25th 2007
               to = "2026-06-21") #at time of download, update as needed
```

```{r}
#preview data
head(optt)
```

```{r}
#summary table
summary(optt)
```

```{r}
#tidy datatable

#rename variables to be more 
optt_tidy <- optt |> 
  rename(open_price = open,
         high_price = high,
         low_price = low,
         close_price = close,
         Volume = volume,
         Adjusted = adjusted) |> 

#break date into day, month, year columns
  separate(date, into = c("Year", "Month", "Day"), sep = "-", convert = TRUE) |> 
  
#add month name column
  mutate(month_name = month.name[Month]) |> 
  
#place month name column after month number
  relocate(month_name, .after = Month) |> 

#create new variables
  mutate(
    #intra-day volatility
    idv = (high_price-low_price),
    #previous close price
    prev_close = lag(close_price),
    #absolute maximum upward movement
    max_up = (abs(high_price - prev_close)),
    #absolute maximum downward movement
    max_down = (abs(low_price - prev_close)),
    #true range:
    #use pmax (parallel max) to take max for each day --> eliminates need for if/else + max()
    true_range = pmax(idv, max_up, max_down, na.rm = TRUE),
    #average true range:
    #use rollmean (zoo package) to look at a sliding window
    ATR = rollmean(true_range, k = 14, fill = NA, align = "right")
  )
```

```{r}
#preview tidy data
head(optt_tidy)
```

```{r}
#view summary of tidy data
summary(optt_tidy)
```

```{r}
#| output: true

#professional table output using kable:
optt_tidy |> 
  select(Year,Month,Day,open_price,close_price,true_range,ATR) |> 
  slice(14:19) |> 
  kable()
```

## Sentiment Analysis

```{r}
#finnhub Sentiment News Scoring

#connect to finnhub
finnhub_url <- paste0(
  "https://finnhub.io/api/v1/company-news?symbol=OPTT&from=2026-05-21&to=2026-06-21&token=", finnhub_key)

#process data using httr and jsonlite
finnhub_raw_data <- GET(finnhub_url)
finnhub_parsed <- fromJSON(content(finnhub_raw_data, "text"))

#preview parsed data
head(finnhub_parsed)
str(finnhub_parsed)

#clean parsed data into a usable format
finnhub_tidy <- finnhub_parsed |> 
  mutate(
    Date = as_date(as_datetime(datetime)),
    Text = headline
  ) |> 
  rename(Source = source) |> 
  select(Date, Text, Source, url)

#only 1 article
```

```{r}
# News API Sentiment Scoring 

#connect to NEWSAPI using newsanchor
news_api_result <- get_everything(
  query = '"Ocean Power Technologies" OR "OPTT stock" OR "NASDAQ:OPTT"',
  from = "2026-05-21",
  to = "2026-06-21",
  api_key = news_api_key
)

#check to see if we got data
news_api_result$metadata$total_results   
news_api_result$metadata$status_code     
news_api_result$results_df  

#3 articles
```

-placeholder for data table with all 4 articles represented and brief discussion of them-

After querying both NewsAPI and Finnhub, we returned a total of 4 relevant articles. 
Due to concerns of limited data and subsequent biases, I will not be using sentiment analysis 
scores within my models. 

Instead, the analysis will focus on price-derived technical indicators alongside trading volume. 
These variables are derived directly from empirical trading data, avoiding the sample-size and 
bias concerns associated with the limited sentiment data.

In order to maintain structural robustness via alternative data streams, the analysis 
will pivot to integrate macroeconomic corporate health metrics, specifically metrics
such as outstanding share dilution and order backlog growth.

## Alternate Data Procurement

-placeholder-

tq_get --> financials --> filter for type == income, unnest quarterly. 

problem --> quarterly means 4 data points per year, not a lot of info (data-sparcity bias)

solution --> forward stepwise function to fill for weekly / monthly data (outstanding shares)

solution --> linear interpolation to fill for weekly / monthly data (order backlog growth)













